---
title: "Hierarchical modeling course notes"
date: "December 13, 2015"
output:
  pdf_document:
    fig_caption: yes
    highlight: tango
    toc: yes
    toc_depth: 3
---

```{r, echo=FALSE, results='hide', message=FALSE}
library(rstan)
rstan_options(auto_write = TRUE)
```

# Preface

This document was made to complement a graduate level course in hierarchical Bayesian models aimed at biology graduate students who have some familiarity with R and statistics. 
While there are many great textbooks out there to go along with such a class, it seemed potentially useful to compile some concise notes with the code and concepts together.
These notes are intended to be used along with other resources, which are listed at the end of each chapter. 
Many people have contributed to the development of these notes including Max Joseph, Will Stutz, Tim Szewczyk, Helen McCreery, Topher Weiss-Lehman, Lauren Shoemaker, and Piet Johnson.
All source code to compile this document is available on GitHub at [https://github.com/hmods/notes](https://github.com/hmods/notes). 

\newpage

```{r child = 'ch1/chapter1.Rmd'}
```

\newpage

```{r child = 'ch2/chapter2.Rmd'}
```

\newpage

```{r child = 'ch3/chapter3.Rmd'}
```

\newpage

```{r child = 'ch4/chapter4.Rmd'}
```

\newpage

```{r child = 'ch5/chapter5.Rmd'}
```

\newpage

```{r child = 'ch6/chapter6.Rmd'}
```

\newpage

```{r child = 'ch7/chapter7.Rmd'}
```

\newpage

```{r child = 'ch8/chapter8.Rmd'}
```
